---
title: "Comparing human and machine translation scores"
author: "Kyo Gerrits"
date: "Generation date: `r format(Sys.time(), '%b %d, %Y - %H:%M:%S')`"
output: 
  html_document:
    theme: sandstone
    highlight: pygments
    toc: true
    df_print: paged
    code_folding: show
    toc_float: 
      collapsed: false
      smooth_scroll: true
    number_sections: true
---
# Loading libraries
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(ggplot2)
library(car)
library(ARTool)
library(glmmTMB)
library(DHARMa)
library(gridExtra)
library(tidyr)
library(dplyr)
library(psych)
library(tibble)
library(purrr)
library(irr)
library(ggpubr)
library(cocor)
library(cluster)
```

# Human annotations
## Loading dataframe
```{r}
df <- read.csv2("Scores_metrics_and_human_creativity.csv", stringsAsFactors = TRUE)
```


## Exploratory data analysis
```{r}
cs_plot <- ggplot(df, aes(x = Modality, y = CS)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total CSs per Modality", x = "Modality", y = "Total CSs")

errors_plot <- ggplot(df, aes(x = Modality, y = ErrorPoints)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(title = "Total Errors per Modality", x = "Modality", y = "Total Errors")

kudos_plot <- ggplot(df, aes(x = Modality, y = Kudos)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  labs(title = "Total Points per Modality", x = "Modality", y = "Total Points")

ci_plot <- ggplot(df, aes(x = Modality, y = C.I.)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Total Points per Modality", x = "Modality", y = "Total Points")

grid.arrange(cs_plot, errors_plot, kudos_plot, ci_plot, ncol=2)
```


## Analyse relations between human annotations and text variables
### First per modality
```{r}
plot_CS_Mod <-ggplot(df, aes(x = Modality, y = CS)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
  theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

plot_ER_Mod <- ggplot(df, aes(x = Modality, y = ErrorPoints)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) +
  theme_bw() +
  theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

plot_CI_Mod <- ggplot(df, aes(x = Modality, y = C.I.)) +
  geom_boxplot(fill = "darkseagreen1") +
  geom_jitter(width = 0.1)+ theme_bw()+
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


```

```{r}
kruskal.test(CS ~ Modality, data = df)
kruskal.test(ErrorPoints ~ Modality, data = df)
kruskal.test(Kudos ~ Modality, data = df)
kruskal.test(C.I. ~ Modality, data = df)
```
#### Post-hoc tests signifiant results


```{r}
pairwise.wilcox.test(df$CS, df$Modality, p.adjust.method = "holm")
pairwise.wilcox.test(df$ErrorPoints, df$Modality, p.adjust.method = "holm")
pairwise.wilcox.test(df$Kudos, df$Modality, p.adjust.method = "holm")
pairwise.wilcox.test(df$C.I., df$Modality, p.adjust.method = "holm")

```


### Then per source language

```{r}
plot_CS_ST <- ggplot(df, aes(x = ST, y = CS)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

plot_ER_ST <- ggplot(df, aes(x = ST, y = ErrorPoints)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_CI_ST <- ggplot(df, aes(x = ST, y = C.I.)) +
  geom_boxplot(fill = "darkseagreen1") +
  geom_jitter(width = 0.1)+  theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

```

```{r}
wilcox.test(CS ~ ST, data = df)
wilcox.test(ErrorPoints ~ ST, data = df)
wilcox.test(Kudos ~ ST, data = df)
wilcox.test(C.I. ~ ST, data = df)
```


### Thirdly, per target language
```{r}
plot_CS_TT <- ggplot(df, aes(x = TT, y = CS)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_ER_TT <- ggplot(df, aes(x = TT, y = ErrorPoints)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_CI_TT <- ggplot(df, aes(x = TT, y = C.I.)) +
  geom_boxplot(fill = "darkseagreen1") +
  geom_jitter(width = 0.1) +theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

```

```{r}
wilcox.test(CS ~ TT, data = df)
wilcox.test(ErrorPoints ~ TT, data = df)
wilcox.test(Kudos ~ TT, data = df)
wilcox.test(C.I. ~ TT, data = df)
```
### Fourthly, per language combination
```{r}
plot_CS_LP<- ggplot(df, aes(x = Language_pair, y = CS)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_ER_LP <- ggplot(df, aes(x = Language_pair, y = ErrorPoints)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)



plot_CI_LP <- ggplot(df, aes(x = Language_pair, y = C.I.)) +
  geom_boxplot(fill = "darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

```

```{r}
kruskal.test(CS ~ Language_pair, data = df)
kruskal.test(ErrorPoints ~ Language_pair, data = df)
kruskal.test(Kudos ~ Language_pair, data = df)
kruskal.test(C.I. ~ Language_pair, data = df)
```

### Lastly, per genre
```{r}
plot_CS_genre <- ggplot(df, aes(x = Genre, y = CS)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_ER_genre <- ggplot(df, aes(x = Genre, y = ErrorPoints)) +
  geom_boxplot(fill="darkseagreen1") +
  geom_jitter(width = 0.1) + theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


plot_CI_genre <- ggplot(df, aes(x = Genre, y = C.I.)) +
  geom_boxplot(fill = "darkseagreen1") +
  geom_jitter(width = 0.1) +theme_bw() +
    theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)

```

```{r}
kruskal.test(CS ~ Genre, data = df)
kruskal.test(ErrorPoints ~ Genre, data = df)
kruskal.test(Kudos ~ Genre, data = df)
kruskal.test(C.I. ~ Genre, data = df)
```
```{r}
pairwise.wilcox.test(df$CS, df$Genre, p.adjust.method = "holm")

```
#### Creating visualisation for the plots

```{r}
# Create a blank plot for spacing
blank <- ggplot() + theme_void()

pdf(file="./boxplots_CI_hum.pdf", width = 10, height = 12)
ggarrange(
    ggarrange(plot_CI_Mod, plot_CI_ST, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(plot_CI_TT, plot_CI_LP, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(blank, plot_CI_genre, blank, nrow = 1,
        widths = c(0.5, 1, 0.5), legend = "none"),
    nrow = 3,
    legend = "none"
    
)

pdf(file="./boxplots_new_mod.pdf", width = 10, height = 7)
ggarrange(
    ggarrange(plot_ER_Mod, plot_CS_Mod, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(blank, plot_CI_Mod, blank, nrow = 1,
        widths = c(0.5, 1, 0.5), legend = "none"),
    nrow = 2,
    legend = "none"
    
)
```

```{r}
pdf(file="./boxplots_new_genre.pdf", width = 10, height = 7)
ggarrange(
    ggarrange(plot_ER_genre, plot_CS_genre, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(blank, plot_CI_genre, blank, nrow = 1,
        widths = c(0.5, 1, 0.5), legend = "none"),
    nrow = 2,
    legend = "none"
    
)
```

```{r}

pdf(file="./boxplots_CS_hum_new.pdf", width = 10, height = 12)
ggarrange(
    ggarrange(plot_CS_Mod, plot_CS_ST, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(plot_CS_TT, plot_CS_LP, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(blank, plot_CS_genre, blank, nrow = 1,
        widths = c(0.5, 1, 0.5), legend = "none"),
    nrow = 3,
    legend = "none"
)
```

```{r}

pdf(file="./boxplots_WE_hum_new.pdf", width = 10, height = 12)

ggarrange(
    ggarrange(plot_ER_Mod, plot_ER_ST, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(plot_ER_TT, plot_ER_LP, blank, nrow = 1,
        widths = c(1, 1, 0), legend = "none"),
    ggarrange(blank, plot_ER_genre, blank, nrow = 1,
        widths = c(0.5, 1, 0.5), legend = "none"),
    nrow = 3,
    legend = "none"
)

```



# Analyse correlation between human and computer scores
```{r}
df$ErrorPoints_rev <- -df$ErrorPoints
df$TER_rev <- -df$TER
df$MetricX24_rev <- -df$MetricX24
human_cols  <- c("CS", "ErrorPoints_rev", "C.I.")
system_cols <- c("BERT.score", "BLEURT", "COMET", "BLEU", "chrF2", "TER_rev", "COMETKiwi", "LiTransProQA", "MetricX24_rev")
df
```
```{r}
# List of metrics
metric_cols <- c("BERT.score", "BLEURT", "COMET", "BLEU", "chrF2",
                 "TER", "COMETKiwi", "LiTransProQA", "MetricX24")

# Convert to long format for the bar metrics
df_long <- df %>%
  pivot_longer(
    cols = all_of(metric_cols),
    names_to = "Metric",
    values_to = "Value"
  )

# Create a combined x-axis label (ggplot can't cluster across 3 variables otherwise)
df_long <- df_long %>%
  mutate(
    Xgroup = interaction(Modality, Language_pair, Genre, sep = "_")
  )

# Also reshape CI for the line
df_CI <- df %>%
  mutate(
    Xgroup = interaction(Modality, Language_pair, Genre, sep = "_")
  )

```

## Create heatmap for correlation
```{r}
#First create a new dataframe with only MT and PE so there are no missing values
df_HT <- df %>%
  filter(!Modality=="HT")
df_HT <- droplevels(df_HT)

#Create datapoints for the heatmap
cor_mat <- matrix(
  nrow = length(human_cols),
  ncol = length(system_cols),
  dimnames = list(human_cols, system_cols)
)

for (h in human_cols) {
  for (s in system_cols) {
    cor_mat[h, s] <- cor(df_HT[[h]], df_HT[[s]], method = "spearman")
  }
}

cor_df <- cor_mat |>
  as.data.frame() |>
  rownames_to_column(var = "human") |>
  pivot_longer(
    cols = -human,
    names_to = "system",
    values_to = "rho"
  )

pdf(file="./heatmap_new.pdf", width = 12, height = 6)
ggplot(cor_df, aes(x = system, y = human, fill = rho)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "steelblue",
    mid = "white",
    high = "firebrick",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  geom_text(aes(label = round(rho, 2)), size = 8) +
  labs(
    x = "Automatic system",
    y = "Human dimension",
    fill = "Spearman ρ"
  ) +
    scale_y_discrete(labels = c("CI", "CS", "Error")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 20),
     axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.y  = element_text(size = 20),
   legend.title = element_text(size = 20),
  legend.text  = element_text(size = 20),
    strip.text = element_text(size = 20),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
  )


 
```

## Check whether human-computer correlation is higher for certain variables
```{r}
group_vars <- c("Modality", "ST", "TT", "Language_pair", "Genre")  # text variables to explore
human_cols  <- c("CS", "ErrorPoints_rev", "C.I.")
system_cols <- c("BERT.score", "BLEURT", "COMET", "BLEU", "chrF2", "TER_rev", "COMETKiwi", "LiTransProQA", "MetricX24_rev")

results <- map_df(human_cols, function(h) {
  map_df(system_cols, function(s) {
    map_df(group_vars, function(g) {
      df_HT %>%
        group_by(.data[[g]]) %>%
        summarise(
          rho = cor(.data[[h]], .data[[s]], method = "spearman"),
          n = n(),
          .groups = "drop"
        ) %>%
        mutate(human_dimension = h, system = s, group_var = g)
    })
  })
})


results <- results %>%
  rename(group_value = group_var)



range_results <- results %>%
  group_by(human_dimension, system) %>%
  summarise(
    min_rho = min(rho, na.rm = TRUE),
    max_rho = max(rho, na.rm = TRUE),
    mean_rho = mean(rho, na.rm = TRUE),
    .groups = "drop"
  )

#pdf(file="./cor_range_new.pdf", width = 16, height = 9)
ggplot(range_results, aes(x = system)) +
  geom_linerange(aes(ymin = min_rho, ymax = max_rho), size = 1.1, color = "black") +
  geom_point(aes(y = mean_rho), size = 3, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  facet_wrap(~ human_dimension, scales = "free_y") +
  labs(
    title = "Range of Spearman correlations across text groupings",
    y = "Correlation (min–max range)",
    x = "Automatic system"
  ) +
  theme_minimal(base_size = 12) +
theme(
  axis.title.x = element_text(size = 16),
  axis.title.y = element_text(size = 16),
  axis.text.x  = element_text(size = 14),
  axis.text.y  = element_text(size = 14),
  plot.title   = element_text(size = 20, face = "bold")  # ← title size here
)


geom_text(aes(y = min_rho, label = round(min_rho, 2)), 
          nudge_y = -0.03, size = 3)

``` 



```{r}
ggplot(results, aes(x = system, y = rho, fill = system)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "R3") +   # pastel, readable, friendly
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  facet_wrap(~ human_dimension) +
  labs(
    title = "Distribution of correlations across grouping variables",
    x = "Automatic system",
    y = "Spearman correlation"
  ) +
  theme_minimal(base_size = 13)
```

```{R}
cor_by_modality <- df_HT %>%
  split(.$Modality) %>%
  map(~{
    mat <- matrix(nrow = length(human_cols),
                  ncol = length(system_cols),
                  dimnames = list(human_cols, system_cols))
    
    for (h in human_cols) {
      for (s in system_cols) {
        mat[h, s] <- cor(.x[[h]], .x[[s]], method = "spearman")
      }
    }
    
    as.data.frame(mat) |>
      tibble::rownames_to_column("human") |>
      pivot_longer(-human, names_to="system", values_to="rho")
  }) |>
  bind_rows(.id="Modality")

ggplot(cor_by_modality, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~Modality) +
  scale_fill_gradient2(limits=c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{R}
cor_by_genre <- df_HT %>%
  split(.$Genre) %>%
  map(~{
    mat <- matrix(nrow = length(human_cols),
                  ncol = length(system_cols),
                  dimnames = list(human_cols, system_cols))
    
    for (h in human_cols) {
      for (s in system_cols) {
        mat[h, s] <- cor(.x[[h]], .x[[s]], method = "spearman")
      }
    }
    
    as.data.frame(mat) |>
      tibble::rownames_to_column("human") |>
      pivot_longer(-human, names_to="system", values_to="rho")
  }) |>
  bind_rows(.id="Genre")

ggplot(cor_by_genre, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~Genre) +
  scale_fill_gradient2(limits=c(-1,1)) +
  scale_y_discrete(labels = c()) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
  axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
  plot.title   = element_text(size = 20, face = "bold"),  # ← title size here
      strip.text = element_text(size = 14),
   legend.title = element_text(size = 16),
  legend.text  = element_text(size = 14)

    )
```

```{r}

pdf(file="./plot_genre_heat.pdf", width = 10, height = 4)
ggplot(cor_by_genre, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~Genre) +
    scale_fill_gradient2(
    low = "royalblue3",
    mid = "white",
    high = "firebrick4",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  scale_y_discrete(labels = c("CI", "CS", "Error")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text.y  = element_text(size = 18),
    plot.title   = element_text(size = 16, face = "bold"),
    strip.text   = element_text(size = 18),
    legend.title = element_text(size = 20),
    legend.text  = element_text(size = 18)
  )

```
```{R}


cor_by_Language_pair <- df_HT %>%
  split(.$Language_pair) %>%
  map(~{
    mat <- matrix(nrow = length(human_cols),
                  ncol = length(system_cols),
                  dimnames = list(human_cols, system_cols))
    
    for (h in human_cols) {
      for (s in system_cols) {
        mat[h, s] <- cor(.x[[h]], .x[[s]], method = "spearman")
      }
    }
    
    as.data.frame(mat) |>
      tibble::rownames_to_column("human") |>
      pivot_longer(-human, names_to="system", values_to="rho")
  }) |>
  bind_rows(.id="Language_pair")

ggplot(cor_by_Language_pair, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~Language_pair) +
  scale_fill_gradient2(limits=c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
cor_by_ST <- df_HT %>%
  split(.$ST) %>%
  map(~{
    mat <- matrix(nrow = length(human_cols),
                  ncol = length(system_cols),
                  dimnames = list(human_cols, system_cols))
    
    for (h in human_cols) {
      for (s in system_cols) {
        mat[h, s] <- cor(.x[[h]], .x[[s]], method = "spearman")
      }
    }
    
    as.data.frame(mat) |>
      tibble::rownames_to_column("human") |>
      pivot_longer(-human, names_to="system", values_to="rho")
  }) |>
  bind_rows(.id="ST")

ggplot(cor_by_ST, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~ST) +
  scale_fill_gradient2(limits=c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
cor_by_TT <- df_HT %>%
  split(.$TT) %>%
  map(~{
    mat <- matrix(nrow = length(human_cols),
                  ncol = length(system_cols),
                  dimnames = list(human_cols, system_cols))
    
    for (h in human_cols) {
      for (s in system_cols) {
        mat[h, s] <- cor(.x[[h]], .x[[s]], method = "spearman")
      }
    }
    
    as.data.frame(mat) |>
      tibble::rownames_to_column("human") |>
      pivot_longer(-human, names_to="system", values_to="rho")
  }) |>
  bind_rows(.id="TT")

ggplot(cor_by_TT, aes(system, human, fill=rho)) +
  geom_tile() +
  facet_wrap(~TT) +
  scale_fill_gradient2(limits=c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Automatic evaluation prompting
```{r}
ann <- read.csv2("Overview_error_annotations.csv", stringsAsFactors = TRUE) 
```

```{r}
str(ann)
head(ann)
tail(ann)
``` 

```{r}
table(ann$Whose_error)
table(ann$Modality, ann$Whose_error)
table(ann$Modality, ann$Severity)
table(ann$Whose_error, ann$Severity)
table(ann$Whose_error, ann$Category_main)
table(ann$Genre, ann$Whose_error)
table(ann$ST, ann$Whose_error)
table(ann$TT, ann$Whose_error)
```

```{r}
table(ann$Whose_error, ann$Severity)
table(ann$Whose_error, ann$Category_main)
table(ann$Whose_error, ann$Modality)
table(ann$Whose_error, ann$Genre)
chisq.test(table(ann$Whose_error, ann$Severity))
chisq.test(table(ann$Whose_error, ann$Category_main))
chisq.test(table(ann$Whose_error, ann$Modality))
chisq.test(table(ann$Whose_error, ann$Genre))

ann %>% 
  count(Modality, Whose_error) %>%
  tidyr::pivot_wider(names_from = Whose_error, values_from = n)
ann %>% 
  count(Genre, Category_main, Whose_error)
```

```{r}
ann_rel <- ann %>%
  filter(
    Whose_error %in% c("Human", "AE") |
      (Whose_error == "Match" & Category_main_match == "Yes")
  ) %>%
  droplevels()

ann_rel_sec <-   ann %>%
  filter(
    Whose_error %in% c("Human", "AE") |
      (Whose_error == "Match")
  ) %>%
  droplevels()

mod_data <- ann_rel_sec %>%
  count(Whose_error, Modality) %>% 
  pivot_wider(names_from = Whose_error, values_from = n, values_fill = 0) %>%
  pivot_longer(cols = c("Human", "AE", "Match"),
               names_to = "Annotator",
               values_to = "Count")

ggplot(mod_data, aes(x = Annotator, y = Modality, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Error Distribution by Annotator and Category",
       x = "Annotator",
       y = "Error Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



heat_data <- ann_rel %>%
  count(Whose_error, Category_main) %>% 
  pivot_wider(names_from = Whose_error, values_from = n, values_fill = 0) %>%
  pivot_longer(cols = c("Human", "AE", "Match"),
               names_to = "Annotator",
               values_to = "Count")

ggplot(heat_data, aes(x = Annotator, y = Category_main, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Error Distribution by Annotator and Category",
       x = "Annotator",
       y = "Error Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ann_rel_sev <- ann %>%
  filter(
    Whose_error %in% c("Human", "AE") |
      (Whose_error == "Match" & Severity_match == "Yes")
  ) %>%
  droplevels()

heat_prop <- ann_rel %>%
  count(Whose_error, Category_main) %>% 
  group_by(Category_main) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()


ggplot(heat_prop, aes(x = Whose_error, y = Category_main, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection",
       x = "Annotator",
       y = "Error Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ST_prop <- ann_rel_sec %>%
  count(Whose_error, ST) %>%
  group_by(ST) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

ggplot(ST_prop, aes(x = Whose_error, y = ST, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "yellow") +
  labs(title = "Proportion Heatmap of Error Detection across Modalities" ,
       x = "Annotator",
       y = "Modality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mod_prop <- ann_rel_sec %>%
  count(Whose_error, Modality) %>%
  group_by(Modality) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

mod_prop2 <- ann_rel_sec %>%
  count(Whose_error, Modality) %>%
  group_by(Whose_error) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

ggplot(mod_prop, aes(x = Whose_error, y = Modality, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection across Modalities" ,
       x = "Annotator",
       y = "Modality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heat_severity <- ann_rel_sev %>%
  count(Whose_error, Severity) %>%
  group_by(Severity) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()


ggplot(heat_severity, aes(Whose_error, Severity, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  labs(title = "Severity Distribution Across Annotators",
       x = "Annotator",
       y = "Severity") +
  theme_minimal()

heat_genre <- ann %>%
  count(Whose_error, Genre) %>%
  group_by(Whose_error) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

ggplot(heat_genre, aes(Whose_error, Genre, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Annotated Errors by Genre",
       x = "Annotator",
       y = "Genre") +
  theme_minimal()

ann_match <- ann %>% filter(Whose_error == "Match")

heat_agree <- ann_match %>%
  count(Category_main, Category_main_2)

ggplot(heat_agree, aes(x = Category_main_2, 
                       y = Category_main, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Human vs AE Agreement Heatmap (Matched Errors)",
       x = "AE Category",
       y = "Human Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heat_agree_sev <- ann_match %>%
  count(Severity, Severity_2)

ggplot(heat_agree_sev, aes(x = Severity_2, 
                       y = Severity, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Human vs AE Agreement Heatmap (Matched Errors)",
       x = "AE Severity",
       y = "Human severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#pdf(file="./plot_heatmap_errors_subcat.pdf", width = 8, height = 6)
ggplot(heat_prop, aes(x = Whose_error, y = Category_main, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection",
       x = "Annotator",
       y = "Error Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )

#pdf(file="./plot_heatmap_errors_sev.pdf", width = 8, height = 6)
ggplot(heat_severity, aes(Whose_error, Severity, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  scale_x_discrete(labels = c("GPT", "Human", "Match")) +
  labs(title = "Severity Distribution Across Annotators",
       x = "Annotator",
       y = "Severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )

ann_no_match2 <- ann %>%
  filter(!Whose_error=="Match")
ann_no_match <-droplevels(ann_no_match2)

mod_prop3 <-ann_no_match2 %>%
  count(Whose_error, Modality) %>%
  group_by(Whose_error) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

    
#pdf(file="./plot_heatmap_errors_mod.pdf", width = 8, height = 6)
ggplot(mod_prop3, aes(x = Whose_error, y = Modality, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection across Modalities" ,
       x = "Annotator",
       y = "Modality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )

mod_genre3 <-ann_no_match2 %>%
  count(Whose_error, Genre) %>%
  group_by(Whose_error) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

    
#pdf(file="./plot_heatmap_errors_mod.pdf", width = 8, height = 6)
ggplot(mod_genre3, aes(x = Whose_error, y = Genre, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection across Modalities" ,
       x = "Annotator",
       y = "Modality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )

mod_ST3 <-ann_no_match2 %>%
  count(Whose_error, ST) %>%
  group_by(ST) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup()

    
#pdf(file="./plot_heatmap_errors_mod.pdf", width = 8, height = 6)
ggplot(mod_prop3, aes(x = Whose_error, y = Modality, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Proportion Heatmap of Error Detection across Modalities" ,
       x = "Annotator",
       y = "Modality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )
```

```{r}
ann_no_match <- ann %>%
  filter(!Whose_error=="Match")
ann_no_match <-droplevels(ann_no_match)
  
ann_no_match

Mod.cont <- cbind (c (+1/3, -2/3, +1/3), c(+0.5, 0, -.5))   # Catalan: +0.5; Dutch: -0.5
colnames (Mod.cont) <- c ("+HTPE-MT", "+HT-PE")
contrasts (ann_no_match$Modality) <- Mod.cont
contrasts (ann_no_match$Modality)

Sev.cont <- cbind (c(+0.5,-.5))   # Catalan: +0.5; Dutch: -0.5
colnames (Sev.cont) <- c ("+Major, -Minor")
contrasts (ann_no_match$Severity) <- Sev.cont
contrasts (ann_no_match$Severity)

Cat.cont <- cbind (c(+1/3, +1/3, -2/3), c(+0.5, -0.5, 0))   # Catalan: +0.5; Dutch: -0.5
colnames (Cat.cont) <- c ("+AccLing-Style", "+Acc-Ling")
contrasts (ann_no_match$Category_main) <- Cat.cont
contrasts (ann_no_match$Category_main)



Genre.cont <- cbind (c (+1/3, +1/3, -2/3), c(+0.5, -0.5, 0))   # Catalan: +0.5; Dutch: -0.5
colnames (Genre.cont) <- c ("+High-Low", "+Poem-ShortStory")
contrasts (ann_no_match$Genre) <- Genre.cont
contrasts (ann_no_match$Genre)

ann_no_match$Whose_error <- relevel(ann_no_match$Whose_error, ref = "Human")

ann_no_match$Whose_error <- ifelse(ann_no_match$Whose_error == "AE", 1, 0)
```
```{r}
model_GPT <- glm(Whose_error ~ Modality + Genre + ST + TT, family = binomial, data = ann_no_match)
summary(model_GPT)

vif(model_GPT)

```

# Creativity annotations
```{r}
crea_df <- read.csv2("Creativity_annotation_together.csv", stringsAsFactors = TRUE)
str(crea_df)
describe(crea_df)
```

```{r}
table(crea_df$Simplified, crea_df$Modality)
table(crea_df$Simplified.1, crea_df$Modality)
table(crea_df$Simplified.1, crea_df$Genre)
table(crea_df$Simplified, crea_df$Genre)

prop.table(table(crea_df$Simplified))
prop.table(table(crea_df$Simplified.1))

``` 

```{r}
Mod.cont <- cbind (c (+1/3, -2/3, +1/3), c(+0.5, 0, -0.5))   # Catalan: +0.5; Dutch: -0.5
colnames (Mod.cont) <- c ("+HTPE-MT", "+HT-PE")
contrasts (crea_df$Modality) <- Mod.cont
contrasts (crea_df$Modality)


Genre.cont <- cbind (c (+1/3, +1/3, -2/3), c(+0.5, -0.5, 0))   # Catalan: +0.5; Dutch: -0.5
colnames (Genre.cont) <- c ("+High-Low", "+Poem-Thriller")
contrasts (crea_df$Genre) <- Genre.cont
contrasts (crea_df$Genre)

TT.cont <- cbind (c (-0.5, +0.5))   # Catalan: +0.5; Dutch: -0.5
colnames (TT.cont) <- c ("+NL-CA")
contrasts (crea_df$TT) <- TT.cont
contrasts (crea_df$TT)

ST.cont <- cbind (c (+0.5, -0.5))   # English: +0.5; Russian: -0.5
colnames (ST.cont) <- c ("+EN-RU")
contrasts (crea_df$ST) <- ST.cont
contrasts (crea_df$ST)

```
 
```{r}
crea_model <- glm(Match ~ Genre + Modality + ST + TT, family=binomial, data=crea_df)
summary(crea_model)
```

```{r}
crea_df_2only <- crea_df %>% 
  filter(!Simplified.1=="O")
droplevels(crea_df_2only)

crea_df_2only$Simplified.1 <- ifelse(crea_df_2only$Simplified.1 == "CS", 1, 0)

Mod.cont <- cbind (c (+1/3, -2/3, +1/3), c(+0.5, 0, -0.5))   # Catalan: +0.5; Dutch: -0.5
colnames (Mod.cont) <- c ("+HTPE-MT", "+HT-PE")
contrasts (crea_df_2only$Modality) <- Mod.cont
contrasts (crea_df_2only$Modality)


Genre.cont <- cbind (c (+1/3, +1/3, -2/3), c(+0.5, -0.5, 0))   # Catalan: +0.5; Dutch: -0.5
colnames (Genre.cont) <- c ("+High-Low", "+Poem-ShortStory")
contrasts (crea_df_2only$Genre) <- Genre.cont
contrasts (crea_df_2only$Genre)

TT.cont <- cbind (c (-0.5, +0.5))   # Catalan: +0.5; Dutch: -0.5
colnames (TT.cont) <- c ("+NL-CA")
contrasts (crea_df_2only$TT) <- TT.cont
contrasts (crea_df_2only$TT)

ST.cont <- cbind (c (+0.5, -0.5))   # English: +0.5; Russian: -0.5
colnames (ST.cont) <- c ("+EN-RU")
contrasts (crea_df_2only$ST) <- ST.cont
contrasts (crea_df_2only$ST)

crea_model2 <- glm(Simplified.1 ~ Genre + Modality + ST + TT, family=binomial, data=crea_df_2only)
summary(crea_model2)

```


```{r}
# basic model with main effects
m_glm <- glm(Match ~ Modality + ST + TT + Genre, family = binomial, data = crea_df)
summary(m_glm)

```


```{r}
df_CI <- read.csv2("Creativity_scores_comb.csv", stringsAsFactors = TRUE)
str(df_CI)
head(df_CI)
describe(df_CI)
```


```{r}
#pdf(file="./plot_scatterplot_CI_new.pdf", width = 8, height = 6)
ggplot(df_CI, aes(x = CI_GPT, y = CI_Human, color = Modality, shape = Genre)) +
  geom_point(alpha = 1, size = 3) +
  geom_smooth(aes(group = Modality), method = "lm", se = FALSE) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  stat_cor(aes(color = Modality, group = Modality), method = "spearman") +
  stat_cor(aes(x = CI_GPT, y = CI_Human),
           inherit.aes = FALSE,
           color = "black",
           label.x = min(df_CI$CI_GPT),
           label.y = max(df_CI$CI_Human),
           method = "spearman") +
  labs(x = "CI_LLM", y = "CI_professional") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text.y  = element_text(size = 20),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 20),
    legend.title = element_text(size = 20),
    legend.text  = element_text(size = 20)
  )
```


```{r}
ggplot(df_CI, aes(x = CI_GPT, y = CI_Human, color = Modality)) +
  geom_point(alpha = 1) +
  geom_text(aes(label = ID), vjust = -0.5, size = 3) +  # Add labels
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  theme_minimal() +
  labs(title = "CS by Trial and Modality") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 20, face = "bold"),
    strip.text   = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14)
  )
```